\textbf{Supplementary Information}

### 2020-09-14 submission/manuscript.Rmd
Coauthor review.

## Summary
Studies of unculturable microbes often combine methods based on DNA, such as 16S rRNA sequencing and metagenomics, with methods that allow an insight into the metabolic status, such as metaproteomics. To apply these procedures to the microbial community inhabiting the surfaces of marine macrophytes it is advisable to perform, prior to the analysis, a selective DNA and protein isolation so that the host material, present in higher quantities, is not hampering the analysis. Two protocols, for DNA and protein isolation, were adapted for selective extractions of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes: the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. Protocols showed an almost complete removal of the epiphytic community regardless of the sampling season, station, settlement or host species. The obtained DNA was suitable for metagenomic and 16S rRNA sequencing, while isolated proteins could be identified by mass spectrometry showing that protocols can be used in 16S rRNA, metagenomic and metaproteomic analysis. Low presence of host DNA and proteins, observed in isolated samples, indicated a selective nature of protocols. Furthermore, the procedures are based on universally available laboratory chemicals making the protocols widely applicable. Taken together, the adapted protocols ensure an almost complete removal of the macrophyte epiphytic community, are selective for microbes inhabiting macrophyte surfaces and are providing DNA and proteins applicable in 16S rRNA sequencing, metagenomics and metaproteomics.

@@

An alternative to these procedures is a direct isolation of the targeted material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed and the isolation procedure continues omitting host material contaminations. To our knowledge the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was provided by @Burke2009. In contrast to previously described methods this protocol enables an almost complete removal of the surface community and it was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, thought providing a selective isolation procedure, uses a rapid multienzyme cleaner (3M) that is not available worldwide and without a known composition [@Burke2009]. Also to our knowledge, no selective isolation protocol for proteins from epiphytic communities inhabiting marine macrophytes was established.

@@

To assess the removal efficiency of DNA and protein isolation procedures leaves and thalli were stained with SYBR Green I and examined under a confocal microscope before and after treatments were performed. To test the isolation protocols on different macrophyte species, DNA and proteins were isolated from the surfaces of two macrophytes, the seagrass *C. nodosa* and the macroalga *C. cylindracea*. *C. nodosa* was sampled in a monospecific meadow (Bay of Saline) and *C. cylindracea* invaded meadow (Bay of Funtana), while samples of *C. cylindracea* originated from a monospecific settlement and a *C. cylindracea* invaded *C. nodosa* meadow in the Bay of Funtana. In addition, the sampling was performed in two contrasting seasons, in December 2017 and in June 2018, to test the impact of seasonal conditions on the removal efficiency. Procedures developed for DNA and protein isolation showed an almost complete removal of the surface community of both, *C. nodosa* and *C. cylindracea*. In addition, a similar removal efficiency was observed for communities sampled in December 2017 (\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}). Also, no effect of station, settlement or isolation procedure (DNA or protein) on the removal efficiency was observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

@@

## Discussion
The study of marine macrophyte epiphytic communities using culture independent techniques, such as 16S rRNA analysis, metagenomics and metaproteomics, requires an initial step of biological material isolation. Methods that have been developed for selective isolation of epiphytic biological material could be divided into two groups: (i) procedures involving, prior to extraction, a cell detachment step such as shaking [@Gross2003; @Noges2010], scraping [@Uku2007] and ultrasonication [@Weidner1996; @Cai2014] and (ii) procedures involving a host tissue incubation aiming at direct lysis of epiphytic microbial cells [@Burke2009]. Protocols that include a cell detachment procedure do not provide complete cell removal [@Richter-Heitmann2016; @Cai2014] and in the case of probe ultrasonication tissue disruption can also be observed [@Richter-Heitmann2016]. To avoid these problems a procedure for DNA isolation involving host tissue incubation in the extraction buffer and direct epiphytic cell lysis on the macrophyte surface was developed [@Burke2009]. Though providing an almost complete cell removal and selective DNA isolation this protocol uses in the extraction buffer a rapid multienzyme cleaner (3M) not available worldwide and without a known composition [@Burke2009]. To circumvent these problems we developed and tested two protocols for selective DNA and protein isolation from marine macrophyte epiphytic communities.

@@

Metagenomic analysis reveals the potential metabolic capacity of microbial communities, while to obtain an insight into the metabolic status of uncultivated members, a metaproteomic approach is required [@Saito2019]. The applied protocol for epipyhtic protein isolation followed by a metaproteomic analysis identified between `r format(min(metap_cog_num), big.mark=',')` and `r format(max(metap_cog_num), big.mark=',')` proteins that is higher than previously reported for e.g. soils [@Chourey2010; @Hultman2015], seawater [@Williams2012] and biofilms [@Leary2014]. The functional annotation of identified proteins into COG functional categories showed that the protein isolation procedure can be used to assess the metabolic status of the macrophyte epiphytic community [@Leary2014]. Similar to metagenomes, the number of identified proteins, including their abundances, associated with Streptophyta in *C. nodosa* samples were low indicating that the procedure is selective for epiphytic cell proteins. In addition, a higher number of identified proteins, and especially their abundances, associated with Chlorophyta was observed in *C. cylindracea* samples. The cause of this elevated presence of Chlorophyta associated proteins can be, as in the case of the DNA isolation procedure, explained by the absence of individual cells in this siphonous alga [@Coneva2015].

@@

\newpage
## Experimental procedures
### Sampling
Leaves of *C. nodosa* were sampled in a *C. nodosa* meadow in the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a *C. nodosa* meadow invaded by *C. cylindracea* in the proximity of the village of Funtana (Bay of Funtana; \ang{45;10;39} N, \ang{13;35;42} E). Thalli of *C. cylindracea* were sampled in the same *C. nodosa* invaded meadow in the Bay Funtana and on a locality of only *C. cylindracea* located in the proximity of the invaded meadow. Leaves and thalli for 16S rRNA analysis, metagenomics and metaproteomics were collected in two contrasting seasons, on 4 December 2017 (16S rRNA analysis and metaproteomics), 14 December 2017 (metagenomics) and 18 June 2018 (16S rRNA analysis, metagenomics and metaproteomics). During spring 2018 the *C. nodosa* meadow in the Bay of Saline decayed to an extent that no leaves could be retrieved [@Najdek2020]. In addition, as not enough DNA for both metagenomic and 16S RNA analysis were obtained during the sampling on 4 December 2017, an additional sampling on 14 December 2017 was carried out in the Bay of Funtana. Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with site seawater. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells.

>>>>

## Summary
Studies of unculturable microbes often combine methods based on DNA, such as 16S rRNA sequencing and metagenomics, with methods that allow insight into the metabolic status, such as metaproteomics. To apply these techniques to the microbial community inhabiting the surfaces of marine macrophytes it is advisable to perform, prior to the analysis, a selective DNA and protein isolation so that the host material, present in higher quantities, is not hampering the analysis. Two protocols, for DNA and protein isolation, were adapted for selective extractions of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes: the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. Protocols showed an almost complete removal of the epiphytic community regardless of the sampling season, station, settlement or host species. The obtained DNA was suitable for metagenomic and 16S rRNA sequencing, while isolated proteins could be identified by mass spectrometry, showing that protocols can be used in 16S rRNA, metagenomic and metaproteomic analysis. Low presence of host DNA and proteins, observed in isolated samples, indicated a selective nature of the protocols. Furthermore, the procedures are based on universally available laboratory chemicals making the protocols widely applicable. Taken together, the adapted protocols ensure an almost complete removal of the macrophyte epiphytic community, are selective for microbes inhabiting macrophyte surfaces and are providing DNA and proteins applicable in 16S rRNA sequencing, metagenomics and metaproteomics.

@@

An alternative to these procedures is a direct isolation of the targeted material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed and the isolation procedure continues, omitting host material contaminations. To our knowledge the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was provided by @Burke2009. In contrast to previously described methods this protocol enables an almost complete removal of the surface community and it was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, thought providing a selective isolation procedure, uses a rapid multienzyme cleaner (3M) that is not available worldwide and without a known composition [@Burke2009]. Also to our knowledge, no selective isolation protocol for proteins from epiphytic communities inhabiting marine macrophytes was established.

@@

To assess the removal efficiency of DNA and protein isolation procedures leaves and thalli were examined under a confocal microscope before and after treatments were performed. Developed procedures showed an almost complete removal of the surface community of both, *C. nodosa* and *C. cylindracea*. In addition, a similar removal efficiency was observed for communities sampled in contrasting seasons, December 2017 (\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}). Also, no effect of station, settlement or isolation procedure (DNA or protein) on the removal efficiency was observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

@@

## Discussion
The study of marine macrophyte epiphytic communities using culture independent techniques, such as 16S rRNA analysis, metagenomics and metaproteomics, requires an initial step of biological material isolation. Methods that have been developed for selective isolation of epiphytic biological material can be divided into two groups: (i) procedures involving, prior to extraction, a cell detachment step such as shaking [@Gross2003; @Noges2010], scraping [@Uku2007] and ultrasonication [@Weidner1996; @Cai2014] and (ii) procedures involving a host tissue incubation aiming at direct lysis of epiphytic microbial cells [@Burke2009]. Protocols that include a cell detachment procedure do not provide complete cell removal [@Richter-Heitmann2016; @Cai2014] and in the case of probe ultrasonication tissue disruption can also be observed [@Richter-Heitmann2016]. 
Though, host tissue incubation procedures provide an almost complete cell removal and selective isolation, existing protocols like the one developed for DNA isolation by @Burke2009 use in the extraction buffer a rapid multienzyme cleaner (3M) not available worldwide and without a known composition [@Burke2009]. To circumvent these problems we developed and tested two protocols for selective DNA and protein isolation from marine macrophyte epiphytic communities.

@@

To obtain an insight into the metabolic status of uncultivated members, a metaproteomic approach is required [@Saito2019]. The applied protocol for epipyhtic protein isolation followed by a metaproteomic analysis identified between `r format(min(metap_cog_num), big.mark=',')` and `r format(max(metap_cog_num), big.mark=',')` proteins, which is higher than previously reported for e.g. soils [@Chourey2010; @Hultman2015], seawater [@Williams2012] and biofilms [@Leary2014]. The functional annotation of identified proteins into COG functional categories showed that the protein isolation procedure can be used to assess the metabolic status of the macrophyte epiphytic community [@Leary2014]. Similar to metagenomes, the number of identified proteins, including their abundances, associated with Streptophyta in *C. nodosa* samples were low indicating that the procedure is selective for epiphytic cell proteins. In addition, a higher number of identified proteins, and especially their abundances, associated with Chlorophyta was observed in *C. cylindracea* samples. The cause of this elevated presence of Chlorophyta associated proteins can be, as in the case of the DNA isolation procedure, explained by the absence of individual cells in this siphonous alga [@Coneva2015].

@@

\newpage
## Experimental procedures
### Sampling
Leaves of *C. nodosa* were sampled in a *C. nodosa* meadow in the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a *C. nodosa* meadow invaded by *C. cylindracea* in the proximity of the village of Funtana (Bay of Funtana; \ang{45;10;39} N, \ang{13;35;42} E). Thalli of *C. cylindracea* were sampled in the same *C. nodosa* invaded meadow in the Bay of Funtana and on a locality of only *C. cylindracea* located in the proximity of the invaded meadow. Leaves and thalli for 16S rRNA analysis, metagenomics and metaproteomics were collected in two contrasting seasons, on 4 December 2017 (16S rRNA analysis and metaproteomics), 14 December 2017 (metagenomics) and 18 June 2018 (16S rRNA analysis, metagenomics and metaproteomics). During spring 2018 the *C. nodosa* meadow in the Bay of Saline decayed to an extent that no leaves could be retrieved [@Najdek2020]. In addition, as not enough DNA for both metagenomic and 16S RNA analysis were obtained during the sampling on 4 December 2017, an additional sampling on 14 December 2017 was carried out in the Bay of Funtana. Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with site seawater. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells.

### 2020-09-14 submission/manuscript.tex
Coauthor review.

Studies of unculturable microbes often combine methods based on DNA,
such as 16S rRNA sequencing and metagenomics, with methods that allow an
insight into the metabolic status, such as metaproteomics. To apply
these procedures to the microbial community inhabiting the surfaces of
marine macrophytes it is advisable to perform, prior to the analysis, a
selective DNA and protein isolation so that the host material, present
in higher quantities, is not hampering the analysis. Two protocols, for
DNA and protein isolation, were adapted for selective extractions of DNA
and proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. Protocols showed an almost
complete removal of the epiphytic community regardless of the sampling
season, station, settlement or host species. The obtained DNA was
suitable for metagenomic and 16S rRNA sequencing, while isolated
proteins could be identified by mass spectrometry showing that protocols
can be used in 16S rRNA, metagenomic and metaproteomic analysis. Low
presence of host DNA and proteins, observed in isolated samples,
indicated a selective nature of protocols. Furthermore, the procedures
are based on universally available laboratory chemicals making the
protocols widely applicable. Taken together, the adapted protocols
ensure an almost complete removal of the macrophyte epiphytic community,
are selective for microbes inhabiting macrophyte surfaces and are
providing DNA and proteins applicable in 16S rRNA sequencing,
metagenomics and metaproteomics.

@@

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, thought providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

@@

To assess the removal efficiency of DNA and protein isolation procedures
leaves and thalli were stained with SYBR Green I and examined under a
confocal microscope before and after treatments were performed. To test
the isolation protocols on different macrophyte species, DNA and
proteins were isolated from the surfaces of two macrophytes, the
seagrass \emph{C. nodosa} and the macroalga \emph{C. cylindracea}.
\emph{C. nodosa} was sampled in a monospecific meadow (Bay of Saline)
and \emph{C. cylindracea} invaded meadow (Bay of Funtana), while samples
of \emph{C. cylindracea} originated from a monospecific settlement and a
\emph{C. cylindracea} invaded \emph{C. nodosa} meadow in the Bay of
Funtana. In addition, the sampling was performed in two contrasting
seasons, in December 2017 and in June 2018, to test the impact of
seasonal conditions on the removal efficiency. Procedures developed for
DNA and protein isolation showed an almost complete removal of the
surface community of both, \emph{C. nodosa} and \emph{C. cylindracea}.
In addition, a similar removal efficiency was observed for communities
sampled in December 2017 (\autoref{confocal_december}) and June 2018
(\autoref{confocal_june}). Also, no effect of station, settlement or
isolation procedure (DNA or protein) on the removal efficiency was
observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

@@

The study of marine macrophyte epiphytic communities using culture
independent techniques, such as 16S rRNA analysis, metagenomics and
metaproteomics, requires an initial step of biological material
isolation. Methods that have been developed for selective isolation of
epiphytic biological material could be divided into two groups: (i)
procedures involving, prior to extraction, a cell detachment step such
as shaking (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping (Uku \emph{et al.}, 2007) and ultrasonication (Weidner \emph{et
al.}, 1996; Cai \emph{et al.}, 2014) and (ii) procedures involving a
host tissue incubation aiming at direct lysis of epiphytic microbial
cells (Burke \emph{et al.}, 2009). Protocols that include a cell
detachment procedure do not provide complete cell removal (Cai \emph{et
al.}, 2014; Richter-Heitmann \emph{et al.}, 2016) and in the case of
probe ultrasonication tissue disruption can also be observed
(Richter-Heitmann \emph{et al.}, 2016). To avoid these problems a
procedure for DNA isolation involving host tissue incubation in the
extraction buffer and direct epiphytic cell lysis on the macrophyte
surface was developed (Burke \emph{et al.}, 2009). Though providing an
almost complete cell removal and selective DNA isolation this protocol
uses in the extraction buffer a rapid multienzyme cleaner (3M) not
available worldwide and without a known composition (Burke \emph{et
al.}, 2009). To circumvent these problems we developed and tested two
protocols for selective DNA and protein isolation from marine macrophyte
epiphytic communities.

@@

Metagenomic analysis reveals the potential metabolic capacity of
microbial communities, while to obtain an insight into the metabolic
status of uncultivated members, a metaproteomic approach is required
(Saito \emph{et al.}, 2019). The applied protocol for epipyhtic protein
isolation followed by a metaproteomic analysis identified between 14,219
and 16,449 proteins that is higher than previously reported for
e.g.~soils (Chourey \emph{et al.}, 2010; Hultman \emph{et al.}, 2015),
seawater (Williams \emph{et al.}, 2012) and biofilms (Leary \emph{et
al.}, 2014). The functional annotation of identified proteins into COG
functional categories showed that the protein isolation procedure can be
used to assess the metabolic status of the macrophyte epiphytic
community (Leary \emph{et al.}, 2014). Similar to metagenomes, the
number of identified proteins, including their abundances, associated
with Streptophyta in \emph{C. nodosa} samples were low indicating that
the procedure is selective for epiphytic cell proteins. In addition, a
higher number of identified proteins, and especially their abundances,
associated with Chlorophyta was observed in \emph{C. cylindracea}
samples. The cause of this elevated presence of Chlorophyta associated
proteins can be, as in the case of the DNA isolation procedure,
explained by the absence of individual cells in this siphonous alga
(Coneva and Chitwood, 2015).

@@

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a \emph{C.
nodosa} meadow invaded by \emph{C. cylindracea} in the proximity of the
village of Funtana (Bay of Funtana; \ang{45;10;39} N, \ang{13;35;42} E).
Thalli of \emph{C. cylindracea} were sampled in the same \emph{C.
nodosa} invaded meadow in the Bay Funtana and on a locality of only
\emph{C. cylindracea} located in the proximity of the invaded meadow.
Leaves and thalli for 16S rRNA analysis, metagenomics and metaproteomics
were collected in two contrasting seasons, on 4 December 2017 (16S rRNA
analysis and metaproteomics), 14 December 2017 (metagenomics) and 18
June 2018 (16S rRNA analysis, metagenomics and metaproteomics). During
spring 2018 the \emph{C. nodosa} meadow in the Bay of Saline decayed to
an extent that no leaves could be retrieved (Najdek \emph{et al.},
2020). In addition, as not enough DNA for both metagenomic and 16S RNA
analysis were obtained during the sampling on 4 December 2017, an
additional sampling on 14 December 2017 was carried out in the Bay of
Funtana. Leaves and thalli were collected by diving and transported to
the laboratory in containers placed on ice and filled with site
seawater. Upon arrival to the laboratory, \emph{C. nodosa} leaves were
cut into sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli
were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were
washed three times with sterile artificial seawater (ASW) to remove
loosely attached microbial cells.

@@

\leavevmode\hypertarget{ref-Bengtsson2020}{}%
Bengtsson, H. (2020) MatrixStats: Functions that apply to rows and
columns of matrices (and to vectors).

@@

\leavevmode\hypertarget{ref-Cucio2018}{}%
Cúcio, C., Overmars, L., Engelen, A.H., and Muyzer, G. (2018)
Metagenomic Analysis Shows the Presence of Bacteria Related to
Free-Living Forms of Sulfur-Oxidizing Chemolithoautotrophic Symbionts in
the Rhizosphere of the Seagrass Zostera marina. \emph{Front Mar Sci}
\textbf{5}:

@@

\leavevmode\hypertarget{ref-Saito2019}{}%
Saito, M.A., Bertrand, E.M., Duffy, M.E., Gaylord, D.A., Held, N.A.,
Hervey, W.J., et al. (2019) Progress and Challenges in Ocean
Metaproteomics and Proposed Best Practices for Data Sharing. \emph{J
Proteome Res} \textbf{18}: 1461--1476.

>>>>

Studies of unculturable microbes often combine methods based on DNA,
such as 16S rRNA sequencing and metagenomics, with methods that allow
insight into the metabolic status, such as metaproteomics. To apply
these techniques to the microbial community inhabiting the surfaces of
marine macrophytes it is advisable to perform, prior to the analysis, a
selective DNA and protein isolation so that the host material, present
in higher quantities, is not hampering the analysis. Two protocols, for
DNA and protein isolation, were adapted for selective extractions of DNA
and proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. Protocols showed an almost
complete removal of the epiphytic community regardless of the sampling
season, station, settlement or host species. The obtained DNA was
suitable for metagenomic and 16S rRNA sequencing, while isolated
proteins could be identified by mass spectrometry, showing that
protocols can be used in 16S rRNA, metagenomic and metaproteomic
analysis. Low presence of host DNA and proteins, observed in isolated
samples, indicated a selective nature of the protocols. Furthermore, the
procedures are based on universally available laboratory chemicals
making the protocols widely applicable. Taken together, the adapted
protocols ensure an almost complete removal of the macrophyte epiphytic
community, are selective for microbes inhabiting macrophyte surfaces and
are providing DNA and proteins applicable in 16S rRNA sequencing,
metagenomics and metaproteomics.

@@

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues, omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, thought providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

@@

To assess the removal efficiency of DNA and protein isolation procedures
leaves and thalli were examined under a confocal microscope before and
after treatments were performed. Developed procedures showed an almost
complete removal of the surface community of both, \emph{C. nodosa} and
\emph{C. cylindracea}. In addition, a similar removal efficiency was
observed for communities sampled in contrasting seasons, December 2017
(\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}).
Also, no effect of station, settlement or isolation procedure (DNA or
protein) on the removal efficiency was observed (Figs.
\ref{confocal_december} and \ref{confocal_june}).

@@

The study of marine macrophyte epiphytic communities using culture
independent techniques, such as 16S rRNA analysis, metagenomics and
metaproteomics, requires an initial step of biological material
isolation. Methods that have been developed for selective isolation of
epiphytic biological material can be divided into two groups: (i)
procedures involving, prior to extraction, a cell detachment step such
as shaking (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping (Uku \emph{et al.}, 2007) and ultrasonication (Weidner \emph{et
al.}, 1996; Cai \emph{et al.}, 2014) and (ii) procedures involving a
host tissue incubation aiming at direct lysis of epiphytic microbial
cells (Burke \emph{et al.}, 2009). Protocols that include a cell
detachment procedure do not provide complete cell removal (Cai \emph{et
al.}, 2014; Richter-Heitmann \emph{et al.}, 2016) and in the case of
probe ultrasonication tissue disruption can also be observed
(Richter-Heitmann \emph{et al.}, 2016). Though, host tissue incubation
procedures provide an almost complete cell removal and selective
isolation, existing protocols like the one developed for DNA isolation
by Burke \emph{et al.} (2009) use in the extraction buffer a rapid
multienzyme cleaner (3M) not available worldwide and without a known
composition (Burke \emph{et al.}, 2009). To circumvent these problems we
developed and tested two protocols for selective DNA and protein
isolation from marine macrophyte epiphytic communities.

@@

To obtain an insight into the metabolic status of uncultivated members,
a metaproteomic approach is required (Saito \emph{et al.}, 2019). The
applied protocol for epipyhtic protein isolation followed by a
metaproteomic analysis identified between 14,219 and 16,449 proteins,
which is higher than previously reported for e.g.~soils (Chourey
\emph{et al.}, 2010; Hultman \emph{et al.}, 2015), seawater (Williams
\emph{et al.}, 2012) and biofilms (Leary \emph{et al.}, 2014). The
functional annotation of identified proteins into COG functional
categories showed that the protein isolation procedure can be used to
assess the metabolic status of the macrophyte epiphytic community (Leary
\emph{et al.}, 2014). Similar to metagenomes, the number of identified
proteins, including their abundances, associated with Streptophyta in
\emph{C. nodosa} samples were low indicating that the procedure is
selective for epiphytic cell proteins. In addition, a higher number of
identified proteins, and especially their abundances, associated with
Chlorophyta was observed in \emph{C. cylindracea} samples. The cause of
this elevated presence of Chlorophyta associated proteins can be, as in
the case of the DNA isolation procedure, explained by the absence of
individual cells in this siphonous alga (Coneva and Chitwood, 2015).

@@

Leaves of \emph{C. nodosa} were sampled in a \emph{C. nodosa} meadow in
the Bay of Saline (\ang{45;7;5} N, \ang{13;37;20} E) and in a \emph{C.
nodosa} meadow invaded by \emph{C. cylindracea} in the proximity of the
village of Funtana (Bay of Funtana; \ang{45;10;39} N, \ang{13;35;42} E).
Thalli of \emph{C. cylindracea} were sampled in the same \emph{C.
nodosa} invaded meadow in the Bay of Funtana and on a locality of only
\emph{C. cylindracea} located in the proximity of the invaded meadow.
Leaves and thalli for 16S rRNA analysis, metagenomics and metaproteomics
were collected in two contrasting seasons, on 4 December 2017 (16S rRNA
analysis and metaproteomics), 14 December 2017 (metagenomics) and 18
June 2018 (16S rRNA analysis, metagenomics and metaproteomics). During
spring 2018 the \emph{C. nodosa} meadow in the Bay of Saline decayed to
an extent that no leaves could be retrieved (Najdek \emph{et al.},
2020). In addition, as not enough DNA for both metagenomic and 16S RNA
analysis were obtained during the sampling on 4 December 2017, an
additional sampling on 14 December 2017 was carried out in the Bay of
Funtana. Leaves and thalli were collected by diving and transported to
the laboratory in containers placed on ice and filled with site
seawater. Upon arrival to the laboratory, \emph{C. nodosa} leaves were
cut into sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli
were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were
washed three times with sterile artificial seawater (ASW) to remove
loosely attached microbial cells.

@@

\leavevmode\hypertarget{ref-Bengtsson2020}{}%
Bengtsson, H. (2020) matrixStats: Functions that apply to rows and
columns of matrices (and to vectors).

@@

\leavevmode\hypertarget{ref-Cucio2018}{}%
Cúcio, C., Overmars, L., Engelen, A.H., and Muyzer, G. (2018)
Metagenomic analysis shows the presence of bacteria related to
free-living forms of sulfur-oxidizing chemolithoautotrophic symbionts in
the rhizosphere of the seagrass \emph{Zostera marina}. \emph{Front Mar
Sci} \textbf{5}: 171.

@@

\leavevmode\hypertarget{ref-Saito2019}{}%
Saito, M.A., Bertrand, E.M., Duffy, M.E., Gaylord, D.A., Held, N.A.,
Hervey, W.J., et al. (2019) Progress and challenges in ocean
metaproteomics and proposed best practices for data sharing. \emph{J
Proteome Res} \textbf{18}: 1461--1476.

### 2020-09-14 submission/supplementary.Rmd
Coauthor review.

```{r out.width="80%", fig.align="center", fig.cap="Rarefaction curves of bacterial and archaeal communities from the surfaces of macrophytes sampled in the Bay of Saline (\\textit{C. nodosa} [Monospecific Settlement]) and the Bay of Funatana (\\textit{C. nodosa} [Mixed Settlement] and \\textit{C. cylindracea} [Mixed and Monospecific Settlement]) in two contrasting seasons.\\label{rarefaction}", fig.pos="H"}
include_graphics("../results/figures/rarefaction.jpg")
```

>>>>

```{r out.width="80%", fig.align="center", fig.cap="Rarefaction curves of bacterial and archaeal communities from the surfaces of two marine macrophytes (\\textit{C. nodosa} and \\textit{C. cylindracea}) sampled in the Bay of Saline and the Bay of Funtana (Mixed and Monospecific Settlements) in two contrasting seasons (4 December 2017 and 19 June 2018).\\label{rarefaction}", fig.pos="H"}
include_graphics("../results/figures/rarefaction.jpg")
```

### 2020-09-14 submission/supplementary.tex
Coauthor review.

\caption{Rarefaction curves of bacterial and archaeal communities from the surfaces of macrophytes sampled in the Bay of Saline (\textit{C. nodosa} [Monospecific Settlement]) and the Bay of Funatana (\textit{C. nodosa} [Mixed Settlement] and \textit{C. cylindracea} [Mixed and Monospecific Settlement]) in two contrasting seasons.\label{rarefaction}}\label{fig:unnamed-chunk-1}
\end{figure}

>>>>

\caption{Rarefaction curves of bacterial and archaeal communities from the surfaces of two marine macrophytes (\textit{C. nodosa} and \textit{C. cylindracea}) sampled in the Bay of Saline and the Bay of Funtana (Mixed and Monospecific Settlements) in two contrasting seasons (4 December 2017 and 19 June 2018).\label{rarefaction}}\label{fig:unnamed-chunk-1}
\end{figure}
